Supp mat: time series paper

### 2017-06-29 writing/supp_mat_draft.tex
added sensitivity analysis to alpha and beta

This supplementary material includes:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Detailed example of subsampling and power calculations
\item
  Additional figures
\item
  Figures and results from other questions of interest
\item
  Additional references
\end{enumerate}

@@

\subsection{Detailed example of subsampling and power
calculations}\label{detailed-example-of-subsampling-and-power-calculations}

@@

Then, we determine an appropriate level of statistical power that we
find acceptable. Traditionally, this has been at 0.8, however, this is
purely historical. Statisical power of 0.8 implies\ldots{}

@@

\subsection{Additional results from the main
manuscript}\label{additional-results-from-the-main-manuscript}

We then correlated minimum time required for each population with its
corresponding life history characteristics. In figure
\ref{fig:biological_correlates} we examined miminmum time required
versus generation length (years), litter size (n), adult body mass
(grams), maximum longevity (years), egg mass (grams), and incubation
(days). None of these variables had much explanatory power in accounting
for the variance in the minimum time required. see table
\ref{table:model_output}\}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

@@

\subsection{Results from other questions of
interest}\label{results-from-other-questions-of-interest}

\subsubsection{Simulations with more complicated population
model}\label{simulations-with-more-complicated-population-model}

In the main text, we showed how a simple population model could be
simulated repeatidly to estimate the power obtained with time series of
increasing length. The model in the main text simulated linear
population growth with only a slope coefficient, y-intercept, and noise
parameter required. This model is purely phenemilogical and does not
include any species life history. Here, we use the same routine as the
main text, but simulate from a more biologically-realistic population
model, the (MODEL NAME HERE):

@@

We simulate this model repetively for different lengths of time series.
Statistical power is the fraction of time series of a set length that
showed signicant trends in abundance, given a true trend in abundance.
\emph{How do I determine significance here? Through linear regression
again, or do I look at non-zero estimation of the parameter values?}

\subsubsection{Empirical approach with more complicated population model
(use model similar to de Valpine
work?)}\label{empirical-approach-with-more-complicated-population-model-use-model-similar-to-de-valpine-work}

\subsubsection{Growth rate calculations}\label{growth-rate-calculations}

Instead of detecting a trend over time with linear regression, we could
also calculate the geometric growth rate of the population. In figure
\ref{fig:growth_rate}, we show how to calculate growth rates for
subsamples of a time series. First, we created subsamples of each
possible length from the full 35 year time series, as we did in the main
next. Next, we calculated the mean and standard deviation of growth
rates for each possible time series length (Fig.
\ref{fig:growth_rate}b). Lastly, we calculated the percent error
\(\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|\)
between the mean of each time series length (observed) and the overall
population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c),
we show the percent error as a function of time series length. Here we
define the minimum time required as the minimum number of years to
achieve less than 20\% error.

\begin{verbatim}
## Loading required package: plyr
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Example of calculating minimum time required for growth rate
estimation. (a) abundance over time, (b) mean and standard deviation of
growth rate for subsamples of entire time series, and (c) the percent
error between mean estimated growth rate and the true longterm growth
rate. The vertical bar denotes the minimum time required to estimate
growth rate within 10\% error.\label{fig:growth_rate}}
\end{figure}

@@

\subsubsection{Using Generalized additive model to identify significant
trends}\label{using-generalized-additive-model-to-identify-significant-trends}

In the main text, we examined the minimum time required to identify a
trend in abundance via linear regression. This approach allowed us to
identify increases or decreases, but a linear model may not always be a
good fit. Generalized additive models (GAMs) are more general than
general linear models and allow more flexible. GAMs are models where a
response variable depends on unknown smooth functions of explanatory
variables. GAMs, therefore, can identify relationships between response
variables and explantory variables that are non-linear and perhaps more
complicated. The downside of GAMs is they typically require more data
and are also prone to overfitting (citations).

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{results here from GAM.\label{fig:gam_example}}
\end{figure}

@@

\subsubsection{Minimum time to detect density-dependence in time
series}\label{minimum-time-to-detect-density-dependence-in-time-series}

For decades, the importance of density-dependence in natural populations
has been an important question in ecology (citations). To address this
question, \citet{Knape2012} used data from the Global Population
Dynamics Database \citep{GPDD2010}. They fitted 627 time series with a
Gompertz population model and found evidence for density-dependence in
45\% of the time series. Each time series in their database contained at
least 15 datapoints, and ranged from 15 to 157. A follow up question to
their work would be, what time series length is required to confidently
estimate density-dependence.

\clearpage
\#\#\# References

\bibliography{White_bib.bib}

>>>>

\tableofcontents

@@

\section{Detailed example of subsampling and power
calculations}\label{detailed-example-of-subsampling-and-power-calculations}

@@

Then, we determine an appropriate level of statistical power that we
find acceptable. Traditionally, this has been at 0.8, however, this is
purely historical. Statistical power of 0.8 implies\ldots{}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{(a) Population size of Harp Seals over time. The line is the
best fit line from linear regresssion. (b) Statistical power for
different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

In Figure \ref{fig:empirical_approach_example}b, statistical power
increases with more time sampled until our desired power of 0.8 at 16
years. Therefore, 16 years would be our estimate for \(T_{min}\).

\section{Additional results from the main
manuscript}\label{additional-results-from-the-main-manuscript}

\subsection{Statistical model of minimum time required and time series
characteristics}\label{statistical-model-of-minimum-time-required-and-time-series-characteristics}

@@

\subsection{Minimum time required and biological
correlates}\label{minimum-time-required-and-biological-correlates}

We then correlated minimum time required for each population with its
corresponding life history characteristics. In figure
\ref{fig:biological_correlates} we examined minimum time required versus
generation length (years), litter size (n), adult body mass (grams),
maximum longevity (years), egg mass (grams), and incubation (days). None
of these variables had much explanatory power in accounting for the
variance in the minimum time required. see table
\ref{table:model_output}\}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

\subsection{Minimum time series and biological
class}\label{minimum-time-series-and-biological-class}

When we examined differences in the minimum time required there were
similar patterns with biological classes. For instance, more time is
required for species within the Actinopterygii class compared to other
species (Fig. \ref{fig:class}a). These differences between biological
classes can be explained by differences in population variability, with
species in Actinopterygii have larger variability in population size
from year-to-year.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{(a) Minimum time required to estimate change in abundance for
species class, (b) longterm trend (estimated slope coefficient) by
species class, (c) interannual variability in population size by species
class, and (d) temporal autocorrelation by species
class.\label{fig:class}}
\end{figure}

\subsection{\texorpdfstring{Sensitivity analysis of \(\alpha\) and
power}{Sensitivity analysis of \textbackslash{}alpha and power}}\label{sensitivity-analysis-of-alpha-and-power}

Estimates of \(T_{min}\) depend strongly on the values used for the
significance level (\(\alpha\)) and the probability of type II error
(\(\beta\)), both of which are set by the practitioner. Here we explore
how estimates of \(T_{min}\) are affected by changes in each of these
parameters. We see that the minimum time required increases with
increases in statistical power or decreases with increases in the
threshold for statistical significance (Fig
\ref{fig:min_time_vs_alpha_beta}).

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{Minimum time required to assess longterm trends in abundance
for values of statistical significance (\(\alpha\)) and power
(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
\end{figure}

\section{Results from other questions of
interest}\label{results-from-other-questions-of-interest}

\subsection{Simulations with more complicated population
model}\label{simulations-with-more-complicated-population-model}

In the main text, we showed how a simple population model could be
simulated repeatedly to estimate the power obtained with time series of
increasing length. The model in the main text simulated linear
population growth with only a slope coefficient, y-intercept, and noise
parameter required. This model is purely phenomenological and does not
include any species life history. Here, we use the same routine as the
main text, but simulate from a more biologically-realistic population
model, the (MODEL NAME HERE):

@@

We simulate this model repetitively for different lengths of time
series. Statistical power is the fraction of time series of a set length
that showed significant trends in abundance, given a true trend in
abundance. \emph{How do I determine significance here? Through linear
regression again, or do I look at non-zero estimation of the parameter
values?}

\subsection{Empirical approach with more complicated population model
(use model similar to de Valpine
work?)}\label{empirical-approach-with-more-complicated-population-model-use-model-similar-to-de-valpine-work}

\subsection{Growth rate calculations}\label{growth-rate-calculations}

Instead of detecting a trend over time with linear regression, we could
also calculate the geometric growth rate of the population. In figure
\ref{fig:growth_rate}, we show how to calculate growth rates for
subsamples of a time series. First, we created subsamples of each
possible length from the full 35 year time series, as we did in the main
next. Next, we calculated the mean and standard deviation of growth
rates for each possible time series length (Fig.
\ref{fig:growth_rate}b). Lastly, we calculated the percent error
\(\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|\)
between the mean of each time series length (observed) and the overall
population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c),
we show the percent error as a function of time series length. Here we
define the minimum time required as the minimum number of years to
achieve less than 10\% error.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Example of calculating minimum time required for growth rate
estimation. (a) abundance over time, (b) mean and standard deviation of
growth rate for subsamples of entire time series, and (c) the percent
error between mean estimated growth rate and the true longterm growth
rate. The vertical bar denotes the minimum time required to estimate
growth rate within 10\% error.\label{fig:growth_rate}}
\end{figure}

@@

\subsection{Using Generalized additive model to identify significant
trends}\label{using-generalized-additive-model-to-identify-significant-trends}

In the main text, we examined the minimum time required to identify a
trend in abundance via linear regression. This approach allowed us to
identify increases or decreases, but a linear model may not always be a
good fit. Generalized additive models (GAMs) are more general than
general linear models and allow more flexible. GAMs are models where a
response variable depends on unknown smooth functions of explanatory
variables. GAMs, therefore, can identify relationships between response
variables and explanatory variables that are non-linear and perhaps more
complicated. The downside of GAMs is they typically require more data
and are also prone to overfitting (citations).

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{results here from GAM.\label{fig:gam_example}}
\end{figure}

@@

\subsection{Minimum time to detect density-dependence in time
series}\label{minimum-time-to-detect-density-dependence-in-time-series}

For decades, the importance of density-dependence in natural populations
has been an important question in ecology (citations). To address this
question, \citet{Knape2012} used data from the Global Population
Dynamics Database \citep{GPDD2010}. They fitted 627 time series with a
Gompertz population model and found evidence for density-dependence in
45\% of the time series. Each time series in their database contained at
least 15 data points, and ranged from 15 to 157. A follow up question to
their work would be, what time series length is required to confidently
estimate density-dependence.

\renewcommand\refname{References}
\bibliography{White_bib.bib}
